---
title: "IS606 Bayesian (Homework 9)"
author: "Daniel Dittenhafer"
date: "December 6, 2015"
output: pdf_document
---
```{r, echo=FALSE} 
library(ggplot2)
library(knitr)
myTheme <- theme(axis.ticks=element_blank(),  
                  panel.border = element_rect(color="gray", fill=NA), 
                  panel.background=element_rect(fill="#FBFBFB"), 
                  panel.grid.major.y=element_line(color="white", size=0.5), 
                  panel.grid.major.x=element_line(color="white", size=0.5),
                  plot.title=element_text(size="10"))
```

### 2.1 (p31)

#### Model A

The following code defines a function for the Model A probabilities.

```{r}
xSeq <- seq(1, 4)
modelA <- function(x)
{
  return (1 / 4)
}

dfModelA <- data.frame(x=xSeq, Px=modelA(xSeq))
kable(dfModelA, caption="Model A Probabilities")
```

We don't see any bias in the Model A probabilities. They are all equal.

#### Model B

The following code defines a function for the Model B probabilities.

```{r}
modelB <- function(x)
{
  return (x / 10)
}

dfModelB <- data.frame(x=xSeq, Px=modelB(xSeq))
kable(dfModelB, caption="Model B Probabilities")
```

Model B is biased toward the higher numbers. 

#### Model C

The following code defines a function for the Model C probabilities.

```{r}
modelC <- function(x)
{
  return (12 / (25 * x))
}

dfModelC <- data.frame(x=xSeq, Px=modelC(xSeq))
kable(dfModelC, caption="Model C Probabilities")
```

Model C is biased toward the lower numbers, slightly more so than Model B was toward the higher numbers.

### 5.1 (p118)

The following code works through the test for a disease and retest scenario for an individual.

```{r}
# Define marginal probabilities
marginalProbDiseasePresent <- 0.001
marginalProbDiseaseAbsent <- 1 - marginalProbDiseasePresent
# Define composite probabilities
probTestPosGivenDisPresent <- 0.99
probTestPosGivenDisAbsent <- 0.05
# Func of Bayes rule
bayesRule <- function(likelihood, prior, evidence)
{
  return ((likelihood * prior) / sum(evidence))
}
```

First we compute the probability of having the disease given the test is positive.

```{r}
# Denominator in component form
evidence <- c(probTestPosGivenDisPresent * marginalProbDiseasePresent,
              probTestPosGivenDisAbsent * marginalProbDiseaseAbsent)
# Call function to compute probability of disease present given a positive test.
probDisPresentGivenTestPost <- bayesRule(probTestPosGivenDisPresent, 
                                         marginalProbDiseasePresent,
                                         evidence)
probDisPresentGivenTestPost
```

Now, after a re-test and a negative result, what is the probability of having the disease?

```{r}
pPrior <- probDisPresentGivenTestPost
probTestNegGivenDisPresent <- (1 - probTestPosGivenDisPresent)
evidence <- c(probTestNegGivenDisPresent * pPrior,
              (1 - probTestPosGivenDisAbsent) * (1 - pPrior))
# Call function to compute probability of disease present given a negative test.
probDisPresentGivenTestNeg <- bayesRule(probTestNegGivenDisPresent, 
                                         pPrior,
                                         evidence)
format(probDisPresentGivenTestNeg, scientific = FALSE)
```

### 5.2 (p118)

#### a) Table

```{r}
n <- 100000
probTestNegGivenDisAbsent <- 1 - probTestPosGivenDisAbsent
dfTable <- data.frame(ThetaSad=c(99, 1, 100), 
                      ThetaHappy=c(probTestPosGivenDisAbsent * marginalProbDiseaseAbsent * n,
                                   probTestNegGivenDisAbsent * marginalProbDiseaseAbsent * n,
                                   99900), 
                      Col3=c(NA, NA, 100000))
dfTable$Col3 <- dfTable$ThetaSad + dfTable$ThetaHappy
dfTable

```

#### b) Proportion of people who have the disease given their test result is positive

It looks like about 2%. Computing:

```{r}
b <- 99 / 5094
b
```


#### c) 

The following code computes the "Markov" natural frequencies and displays the table.

```{r}
nDisPresent <- 10000
nDisAbsent <- 9990000
dfNaturalFreq <- data.frame(ThetaSad=c(nDisPresent, 
                                       nDisPresent * (0.99), 
                                       (nDisPresent * (0.99)) * ((1-0.99))), 
                            ThetaHappy=c(nDisAbsent,
                                         nDisAbsent * (0.05),
                                         (nDisAbsent * (0.05)) * (1-0.05)))
kable(dfNaturalFreq)
```

#### d)

What proportion of people who test positive at first and then negative on retest, actually have the disease?

```{r}
numerator <- ((nDisPresent * (0.99)) * ((1-0.99)))
denom <- ((nDisPresent * (0.99)) * (1-0.99)) + (nDisAbsent * (0.05) * (1-0.05))
nActuallyHaveDisease <- numerator / denom

nActuallyHaveDisease
```