---
title: "IS606 Bayesian (Homework 9)"
author: "Daniel Dittenhafer"
date: "December 6, 2015"
output: pdf_document
---
```{r, echo=FALSE} 
library(ggplot2)
library(knitr)
myTheme <- theme(axis.ticks=element_blank(),  
                  panel.border = element_rect(color="gray", fill=NA), 
                  panel.background=element_rect(fill="#FBFBFB"), 
                  panel.grid.major.y=element_line(color="white", size=0.5), 
                  panel.grid.major.x=element_line(color="white", size=0.5),
                  plot.title=element_text(size="10"))
```

### 2.1 (p31)

#### Model A

The following code defines a function for the Model A probabilities.

```{r}
xSeq <- seq(1, 4)
modelA <- function(x)
{
  return (1 / 4)
}

dfModelA <- data.frame(x=xSeq, Px=modelA(xSeq))
kable(dfModelA, caption="Model A Probabilities")
```

We don't see any bias in the Model A probabilities. They are all equal.

#### Model B

The following code defines a function for the Model B probabilities.

```{r}
modelB <- function(x)
{
  return (x / 10)
}

dfModelB <- data.frame(x=xSeq, Px=modelB(xSeq))
kable(dfModelB, caption="Model B Probabilities")
```

Model B is biased toward the higher numbers. 

#### Model C

The following code defines a function for the Model C probabilities.

```{r}
modelC <- function(x)
{
  return (12 / (25 * x))
}

dfModelC <- data.frame(x=xSeq, Px=modelC(xSeq))
kable(dfModelC, caption="Model C Probabilities")
```

Model C is biased toward the lower numbers, slightly more so than Model B was toward the higher numbers.

### 5.1

### 5.2
