---
title: "IS606 Homework 5"
author: "Daniel Dittenhafer"
date: "October 22, 2015"
output: pdf_document
---
```{r, echo=FALSE} 
library(ggplot2)
myTheme <- theme(axis.ticks=element_blank(),  
                  panel.border = element_rect(color="gray", fill=NA), 
                  panel.background=element_rect(fill="#FBFBFB"), 
                  panel.grid.major.y=element_line(color="white", size=0.5), 
                  panel.grid.major.x=element_line(color="white", size=0.5),
                  plot.title=element_text(size="10"))
```
### 5.6 Working Backwards, Part II (p257)

**A 90% confidence interval for a population mean is (65, 77). The population distribution is approx. normal and the population standard deviation is unknown. This confidence interval is based on a simple random sample of 25 observations. Calculate the sample mean, the margin of error and the sample standard deviation.**


First we can work out the sample mean, by finding the mid point of the confidence interval range. We also get the T * SE segment of the confidence interval equation out of these computations.

```{r}
n <- 25
diff <- 77 - 65
TxSE <- diff / 2

mean56 <- 65 + TxSE
mean56
```

Next we need to resolve the t value for a 90% confidence interval of n=25, df=24:

```{r}
df <- n - 1
t <- qt(.1, df)

t

```

### 5.14

### 5.20

### 5.32

### 5.48
