---
title: "Homework 3"
author: "Daniel Dittenhafer"
date: "September 19, 2015"
output: pdf_document
---
```{r, echo=FALSE} 
library(ggplot2)
myTheme <- theme(axis.ticks=element_blank(),  
                  panel.border = element_rect(color="gray", fill=NA), 
                  panel.background=element_rect(fill="#FBFBFB"), 
                  panel.grid.major.y=element_line(color="white", size=0.5), 
                  panel.grid.major.x=element_line(color="white", size=0.5),
                  plot.title=element_text(size="10"))
```

### 3.2 Area under the curve Part II (p158)

***What percent of a standard normal distribution \(N(\mu = 0, \sigma = 1)\) is found in each region? Be sure to draw a graph.***

#### a. Z > -1.13

First, we define in `R` code the Z score and then use the `pnorm` function to determine the percentage on the left tail. Subtract this value from 1 to find the right tail value.

```{r}
zGt <- -1.13
pGt <- 1 - pnorm(zGt)
pGt
```

The percent of the standard normal distribution found in the region Z > `r zGt` is `r pGt`.

```{r, echo=FALSE}
x <- seq(-4, 4, length=100)
hx <- dnorm(x)
df <- data.frame(x, hx)
dfRegion <- df[df$x > zGt, ]

g1 <- ggplot(data=df) + 
  geom_line(aes(x=x, y=hx)) + 
  geom_ribbon(data=dfRegion, aes(ymin=0, ymax=hx, x=x), fill="lightblue") +
  labs(title="a) Region Greater Than Z = -1.13", y="", x="") +
  myTheme
g1
```

#### b. Z < 0.18

First, we define in `R` code the Z score and then use the `pnorm` function to determine the percentage on the left tail (less than). 

```{r}
zLt <- 0.18
pLt <- pnorm(zLt)
pLt
```

The percent of the standard normal distribution found in the region Z < `r zLt` is `r pLt`.

```{r, echo=FALSE}
x <- seq(-4, 4, length=100)
hx <- dnorm(x)
df <- data.frame(x, hx)
dfRegion <- df[df$x < zLt, ]

g1 <- ggplot(data=df) + 
  geom_line(aes(x=x, y=hx)) + 
  geom_ribbon(data=dfRegion, aes(ymin=0, ymax=hx, x=x), fill="lightblue") +
  labs(title="b) Region Less Than Z = 0.18", y="", x="") +
  myTheme
g1
```

#### c. Z > 8

Again, we define in `R` code the Z score and then use the `pnorm` function to determine the percentage on the right tail (greater than).

```{r}
zGt <- 8
pGt <- 1 - pnorm(zGt)
round(pGt, 4)
```

The percent of the standard normal distribution found in the region Z > `r zGt` is `r round(pGt, 4)`. This particular scenario is so extreme that it doesn't even show on the visualization.

```{r, echo=FALSE}
x <- seq(-9, 9, length=100)
hx <- dnorm(x)
df <- data.frame(x, hx)
dfRegion <- df[df$x > zGt, ]

g1 <- ggplot(data=df) + 
  geom_line(aes(x=x, y=hx)) + 
  geom_ribbon(data=dfRegion, aes(ymin=0, ymax=hx, x=x), fill="lightblue") +
  labs(title="c) Region Greater Than Z = 8", y="", x="") +
  myTheme
g1
```

#### d. |Z| < 0.5

Again, we define in `R` code the Z score and then use the `pnorm` function to determine the percentage on the left tail. Due to the absolute value sign on the Z, we are looking for the middle region and subtract the pnorm value from 0.5. Also, this becomes a two tail-like question and we therefore double the value resulting from the `pnorm` subtraction.

```{r}
zAbs <- 0.5
pAbs <- 2 * (0.5 - pnorm(-1 * zAbs))
round(pAbs, 4)
```

The percent of the standard normal distribution found in the region |Z| < `r zAbs` is `r round(pAbs, 4)`. 

```{r, echo=FALSE}
x <- seq(-4, 4, length=100)
hx <- dnorm(x)
df <- data.frame(x, hx)
dfRegion <- df[(-1 * zAbs) < df$x & df$x < zAbs, ]

g1 <- ggplot(data=df) + 
  geom_line(aes(x=x, y=hx)) + 
  geom_ribbon(data=dfRegion, aes(ymin=0, ymax=hx, x=x), fill="lightblue") +
  labs(title="d) Region |Z| < 0.5", y="", x="") +
  myTheme
g1
```

### 3.4 Triathlon times, Part I (p158)

Racer | Group          | Time (sec) |
------|----------------|------------|
Leo   | Men, 30 - 34   | 4948       |  
Mary  | Women, 25 - 29 | 5513       |

Group        | Mean | Stdev |
-------------|------|-------|
Men, 30-34   | 4313 | 583   |
Women, 25-29 | 5261 | 807   |

* Normally distributed finishing times for both groups

#### a) Write down the short-hand for these two normal distributions.

The short-hand for these two normal distributions follows:

Men, 30-34: \(N(\mu = 4313, \sigma = 583)\)

Women, 25-29: \(N(\mu = 5261, \sigma = 807)\)

#### b) What are the Z-scores for Leo's and Mary's finishing times? What do these Z-score tell you?

Using `R`, we define the mean, standard deviation and individual times. Then we compute the Z score using the equation \(Z = \frac{x-\mu}{\sigma}\) for Leo and Mary.

```{r}
men3034mean <- 4313
men3034sd <- 583
leoTime <- 4948

leoZc <- (leoTime - men3034mean) / men3034sd
leoZc

women3034mean <- 5261
women3034sd <- 807
maryTime <- 5513

maryZc <- (maryTime - women3034mean) / women3034sd
maryZc
```



### 3.18  (p?)

### 3.22  (p?)

### 3.38  (p?)

### 3.42 (p?)