---
title: "IS606 Homework 5"
author: "Daniel Dittenhafer"
date: "October 22, 2015"
output: pdf_document
---
```{r, echo=FALSE} 
library(ggplot2)
myTheme <- theme(axis.ticks=element_blank(),  
                  panel.border = element_rect(color="gray", fill=NA), 
                  panel.background=element_rect(fill="#FBFBFB"), 
                  panel.grid.major.y=element_line(color="white", size=0.5), 
                  panel.grid.major.x=element_line(color="white", size=0.5),
                  plot.title=element_text(size="10"))
```
### 5.6 Working Backwards, Part II (p257)

**A 90% confidence interval for a population mean is (65, 77). The population distribution is approx. normal and the population standard deviation is unknown. This confidence interval is based on a simple random sample of 25 observations. Calculate the sample mean, the margin of error and the sample standard deviation.**


First we can work out the sample mean, by finding the mid point of the confidence interval range. We also get the T * SE segment of the confidence interval equation out of these computations.

```{r}
n <- 25
diff <- 77 - 65
marginOfError <- diff / 2

mean56 <- 65 + marginOfError
mean56
```

Next we need to resolve the t value for a 90% confidence interval of n=25, df=24. We use the `qt` function to look up the t value in the t distribution for the \(1 - \alpha / 2 = 0.95\) where \(\alpha = 1 - CI = 0.10\):

```{r}
df <- n - 1
t <- qt(.95, df)
t
```

With the t value, we can separate the standard error:

```{r}
SE <- marginOfError / t
SE
```

With the standard error and the number of cases we can back into the sample standard deviation:

```{r}
sdSample <- SE * sqrt(n)
sdSample
```

### 5.14 SAT Scores (p259)

***SAT Scores of students at an Ivy League college are distributed with a standard deviation of 250 points. Two statistitcs students, Raina and Luke, want to estimate the average SAT score of students at this college as part of a class project.  They want their margin of error to be no more than 25 points.***

#### a) Raina wants to use a 90% confidence interval. How large of a sample should she collect?

Using the normal distribution Z score, we can compute the sample size needed as follows:

```{r}
mE <- 25
S <- 250
z <- qnorm(0.95)
z
n <- ((S * z) / mE)^2
n
```

#### b) Luke wants to use a 99% confidence interval. Without calculating the actual sample size, determine whether his sample should be larger or smaller than Raina's, and explain your reasoning.

In order to achieve a 99% confidence interval while maintaining the margin of error of 25 points, a larger sample size will be needed sufficient to offset the increased Z score associated with the 99% CI.

#### c) Calculate the minimum required sample size for Luke.

```{r}
z <- qnorm(0.995)
z
n <- ((S * z) / mE)^2
n
```

### 5.20

### 5.32

### 5.48
