---
title: "IS606 - Homework 8"
author: "Daniel Dittenhafer"
date: "December 3, 2015"
output: pdf_document
---
```{r, echo=FALSE} 
library(ggplot2)
myTheme <- theme(axis.ticks=element_blank(),  
                  panel.border = element_rect(color="gray", fill=NA), 
                  panel.background=element_rect(fill="#FBFBFB"), 
                  panel.grid.major.y=element_line(color="white", size=0.5), 
                  panel.grid.major.x=element_line(color="white", size=0.5),
                  plot.title=element_text(size="10"))
```

### 8.2 Baby Weights, Part II (p395)

#### a) Write the equation of the regression line.

\[y=120.07 - 1.93 x_{parity}\]

#### b) Interpret the slope in the context, and calculate the predicted birth weight of first borns and others.

The slope \(120.07\) indicates the first born (parity = 0) would be predicted to weigh 120.07 ounces. The others born, based on the slope of -1.93 would be \(`r 120.07 - 1.93`\) oz.


#### c) Is there a statistically significant relationship between the average birth weight and parity?

Given the p-value of 0.1052 for the parity parameter, I conclude there is not a statistically significant relationship between average birth weight and parity.

### 8.4 Absenteeism (p397)

#### a) Write the equation of the regression line.

\[y = 18.93 - 9.11 x_{eth} + 3.10 x_{sex} + 2.15 x_{lrn}\]


#### b) Interpret each one of the slopes in this context.

* The slope of `eth` indicates that, all else being equal, there is a 9.11 day reduction in the predicted absenteeism when the subject is no aboriginal.

* The slope of `sex` indicates that, all else being equal, there is a 3.10 day increase in the predicted absenteeism when the subject is male.

* The slope of `lrn` indicates that, all else being equal, there is a 2.15 day increase in the predicted absenteeism when the subject is a slow learner.


#### c) Calculate the residual for the first observation in the data set: a student who is aboriginal, male, a slow learner, and missed 2 days of school.

Using the R code below, we compute the predicted absentee days and further compute the residual against the actual missed days of school.

```{r}
eth <- 0
sex <- 1
lrn <- 1
actualDaysMissed <- 2

absDaysPred <- 18.93 - 9.11 * eth + 3.1 * sex + 2.15 * lrn
absDaysPred

residual <- actualDaysMissed - absDaysPred
residual
```

The residual is `r residual` days.

#### d) The variance of the residuals is 240.57 and the variance of the number of absent days for all students in the data set is 264.17. Calculate the \(R^2\) and adjusted \(R^2\). Note that there are 146 observations in the data set.

The following R code computes the \(R^2\) and adjusted \(R^2\):

```{r}
n <- 146
k <- 3
varRes <- 240.57
varOut <- 264.17

R2 <- 1 - (240.57 / 264.17)
R2

adjustedR2 <- R2 * ( (n-1) / (n-k-1) )
adjustedR2

```


### 8.8 Absenteeism, Part II (p399)



### 8.16

### 8.18



