---
title: "3.1 Area Under the Curve"
author: "Daniel Dittenhafer"
date: "September 24, 2015"
output:
  ioslides_presentation:
    smaller: true
---
```{r, echo=FALSE} 
library(ggplot2)
myTheme <- theme(axis.ticks=element_blank(),  
                  panel.border = element_rect(color="gray", fill=NA), 
                  panel.background=element_rect(fill="#FBFBFB"), 
                  panel.grid.major.y=element_line(color="white", size=0.5), 
                  panel.grid.major.x=element_line(color="white", size=0.5),
                  plot.title=element_text(size="10"))
```
<style type="text/css">
    ol { list-style-type: lower-alpha; }
</style>

## 3.1 Area Under the Curve

What percent of a standard normal distribution \(N(\mu = 0, \sigma = 1)\) is found in each region? Be sure to draw a graph.

a. Z < -1.35
b. Z >  1.48
c. -0.4 < Z < 1.5
d. |Z| > 2

## (a) Z < -1.35

First, we define in `R` code the Z score and then use the `pnorm` function to determine the percentage on the left tail. 

```{r}
zLt <- -1.35
pLt <- pnorm(zLt)
pLt
```


## (a) Z < -1.35 Visualization

The percent of the standard normal distribution found in the region Z < `r zLt` is `r round(pLt * 100, 2)`%.

```{r, echo=FALSE}
x <- seq(-4, 4, length=100)
hx <- dnorm(x)
df <- data.frame(x, hx)
dfRegion <- df[df$x < zLt, ]

g1 <- ggplot(data=df) + 
  geom_line(aes(x=x, y=hx)) + 
  geom_ribbon(data=dfRegion, aes(ymin=0, ymax=hx, x=x), fill="lightblue") +
  labs(title="a) Region Less Than Z = -1.35", y="", x="") +
  myTheme
g1
```

## (b) Z >  1.48

Again, we define in `R` code the Z score and then use the `pnorm` function to determine the percentage on the left tail. This time we subtract the value returned from `pnorm` from 1 to convert to the right tail value.

```{r}
zGt <- 1.48
pGt <- 1 - pnorm(zGt)
pGt
```

## (b) Z >  1.48 Visualization

The percent of the standard normal distribution found in the region Z > `r zGt` is `r round(pGt * 100, 2)`%.

```{r, echo=FALSE}
x <- seq(-4, 4, length=100)
hx <- dnorm(x)
df <- data.frame(x, hx)
dfRegion <- df[df$x > zGt, ]

g1 <- ggplot(data=df) + 
  geom_line(aes(x=x, y=hx)) + 
  geom_ribbon(data=dfRegion, aes(ymin=0, ymax=hx, x=x), fill="lightblue") +
  labs(title="b) Region Greater Than Z = 1.48", y="", x="") +
  myTheme
g1
```

## (c) -0.4 < Z < 1.5

Again, we define in `R` code the Z scores and use `pnorm`. This time we have a middle region with a portion below zero and a portion above zero, therefore we substract using 0.5 depending on which side we are on.

```{r}
z1 <- -0.4
p1 <- 0.5 - pnorm(z1) # Left of mean 0

z2 <- 1.5
p2 <- pnorm(z2) - 0.5 # Right of mean 0

pT <- p1 + p2
pT
```

## (c) -0.4 < Z < 1.5 Visualization

The percent of the standard normal distribution found in the region -0.4 < Z < 1.5 is `r round(pT * 100, 2)`%.

```{r, echo=FALSE}
x <- seq(-4, 4, length=100)
hx <- dnorm(x)
df <- data.frame(x, hx)
dfRegion <- df[z1 < df$x & df$x < z2, ]

g1 <- ggplot(data=df) + 
  geom_line(aes(x=x, y=hx)) + 
  geom_ribbon(data=dfRegion, aes(ymin=0, ymax=hx, x=x), fill="lightblue") +
  labs(title="c) Region -0.4 < Z < 1.5", y="", x="") +
  myTheme
g1
```

## (d) Abs(Z) > 2

Once again, we define in `R` code the Z score and use `pnorm`. We have a two separate tail regions with a portion below 0 and a portion above zero (not contiguous). In this case, the regions are identical due to the absolute value, therefore we can simply multiply by 2.

```{r}
z <- 2
p1 <- 1 - pnorm(z) # Right tail

pT <- 2 * p1
pT
```

## (d) Abs(Z) > 2 Visualization

The percent of the standard normal distribution found in the region \( |Z| > 2\) is `r round(pT * 100, 2)`%.

```{r, echo=FALSE}
x <- seq(-4, 4, length=500)
hx <- dnorm(x)
df <- data.frame(x, hx)
dfRegionLeft <- df[df$x < (-1 * z), ]
dfRegionRight <- df[z < df$x, ]

g1 <- ggplot(data=df) + 
  geom_line(aes(x=x, y=hx)) + 
  geom_ribbon(data=dfRegionLeft, aes(ymin=0, ymax=hx, x=x), fill="lightblue") +
  geom_ribbon(data=dfRegionRight, aes(ymin=0, ymax=hx, x=x), fill="lightblue") +
  labs(title="d) Region |Z| > 2", y="", x="") +
  myTheme
g1
```

