---
title: "Multiple linear regression"
output:
  pdf_document: 
    fig_width: 4
    fig_height: 3.5
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
---
```{r, echo=FALSE} 
library(ggplot2)
myTheme <- theme(axis.ticks=element_blank(),  
                  panel.border = element_rect(color="gray", fill=NA), 
                  panel.background=element_rect(fill="#FBFBFB"), 
                  panel.grid.major.y=element_line(color="white", size=0.5), 
                  panel.grid.major.x=element_line(color="white", size=0.5),
                  plot.title=element_text(size="10"))
```

## Grading the professor

Many college courses conclude by giving students the opportunity to evaluate 
the course and the instructor anonymously. However, the use of these student 
evaluations as an indicator of course quality and teaching effectiveness is 
often criticized because these measures may reflect the influence of 
non-teaching related characteristics, such as the physical appearance of the 
instructor. The article titled, "Beauty in the classroom: instructors' 
pulchritude and putative pedagogical productivity" (Hamermesh and Parker, 2005) 
found that instructors who are viewed to be better looking receive higher 
instructional ratings. (Daniel S. Hamermesh, Amy Parker, Beauty in the 
classroom: instructors pulchritude and  putative pedagogical productivity, 
*Economics of Education Review*, Volume 24, Issue 4, August 2005, Pages 369-376, 
ISSN 0272-7757, 10.1016/j.econedurev.2004.07.013. [http://www.sciencedirect.com/science/article/pii/S0272775704001165](http://www.sciencedirect.com/science/article/pii/S0272775704001165).)

In this lab we will analyze the data from this study in order to learn what goes 
into a positive professor evaluation.

## The data

The data were gathered from end of semester student evaluations for a large 
sample of professors from the University of Texas at Austin. In addition, six 
students rated the professors' physical appearance. (This is aslightly modified 
version of the original data set that was released as part of the replication 
data for *Data Analysis Using Regression and Multilevel/Hierarchical Models* 
(Gelman and Hill, 2007).) The result is a data frame where each row contains a 
different course and columns represent variables about the courses and professors.

```{r load-data, eval=TRUE}
load("more/evals.RData")
```

variable         | description
---------------- | -----------
`score`          | average professor evaluation score: (1) very unsatisfactory - (5) excellent.
`rank`           | rank of professor: teaching, tenure track, tenured.
`ethnicity`      | ethnicity of professor: not minority, minority.
`gender`         | gender of professor: female, male.
`language`       | language of school where professor received education: english or non-english.
`age`            | age of professor.
`cls_perc_eval`  | percent of students in class who completed evaluation.
`cls_did_eval`   | number of students in class who completed evaluation.
`cls_students`   | total number of students in class.
`cls_level`      | class level: lower, upper.
`cls_profs`      | number of professors teaching sections in course in sample: single, multiple.
`cls_credits`    | number of credits of class: one credit (lab, PE, etc.), multi credit.
`bty_f1lower`    | beauty rating of professor from lower level female: (1) lowest - (10) highest.
`bty_f1upper`    | beauty rating of professor from upper level female: (1) lowest - (10) highest.
`bty_f2upper`    | beauty rating of professor from second upper level female: (1) lowest - (10) highest.
`bty_m1lower`    | beauty rating of professor from lower level male: (1) lowest - (10) highest.
`bty_m1upper`    | beauty rating of professor from upper level male: (1) lowest - (10) highest.
`bty_m2upper`    | beauty rating of professor from second upper level male: (1) lowest - (10) highest.
`bty_avg`        | average beauty rating of professor.
`pic_outfit`     | outfit of professor in picture: not formal, formal.
`pic_color`      | color of professor's picture: color, black & white.

## Exploring the data

1.  Is this an observational study or an experiment? The original research
    question posed in the paper is whether beauty leads directly to the
    differences in course evaluations. Given the study design, is it possible to
    answer this question as it is phrased? If not, rephrase the question.
    
    **This is an experiment but it uses observational data in both the the treatment (beauty measure), and with the course evaluation (as the response variable). Although the reseach is designed as an experiment, all aspects are extremely subjective any doesn't seem to be a good design for causal inference. A better research question would be whether beauty is correlated with differences in course evaluations.**

2.  Describe the distribution of `score`. Is the distribution skewed? What does 
    that tell you about how students rate courses? Is this what you expected to 
    see? Why, or why not?
    
    **The distribution of `score` is skewed left. This suggests students tend to rate fairly high typically. I had not developed any expectation before running the analysis, but I guess I would have assumed a bit more normalness and less skew in the sense that not all professors are as good as others and this should be reflected in the scores.**
  
```{r}
g1 <- ggplot(data=evals) + 
  geom_histogram(aes(x=score)) + 
  labs(title="Distribution of Eval Score") +
  myTheme
g1
```

3.  Excluding `score`, select two other variables and describe their relationship 
    using an appropriate visualization (scatterplot, side-by-side boxplots, or 
    mosaic plot).
    
```{r}
g1 <- ggplot(data=evals, aes(x=age, y=bty_avg)) + 
  geom_point() + 
  geom_smooth() +
  labs(title="Scatterplot of Age vs Beauty Average") +
  myTheme
g1
```    
    

## Simple linear regression

The fundamental phenomenon suggested by the study is that better looking teachers
are evaluated more favorably. Let's create a scatterplot to see if this appears 
to be the case:

```{r scatter-score-bty_avg, eval = TRUE}
g1 <- ggplot(data=evals) + 
  geom_point(aes(x=bty_avg, y=score), position="dodge") + myTheme
g1

nrow(evals)
```


Before we draw conclusions about the trend, compare the number of observations 
in the data frame with the approximate number of points on the scatterplot. 
Is anything awry?

**Approximate number of points in the scatterplot? That would be an awfully poor estimate if I made one. I dont' see anything awry. Maybe I'm missing something. **

4.  Replot the scatterplot, but this time use the function `jitter()` on the
    $y$- or the $x$-coordinate. (Use `?jitter` to learn more.) What was
    misleading about the initial scatterplot?
    **Ok, so there were duplicate scores that were overlapping in the original plot.**
    
    ```{r}
    plot(jitter(evals$score) ~ jitter(evals$bty_avg))
    
    ```

5.  Let's see if the apparent trend in the plot is something more than
    natural variation. Fit a linear model called `m_bty` to predict average
    professor score by average beauty rating and add the line to your plot
    using `abline(m_bty)`. Write out the equation for the linear model and
    interpret the slope. Is average beauty score a statistically significant
    predictor? Does it appear to be a practically significant predictor?
    
    ```{r}
    m_bty <- lm(evals$score ~ evals$bty_avg)
    plot(jitter(evals$score) ~ jitter(evals$bty_avg))

    abline(m_bty)
    summary(m_bty)
    ```
    
    \[\hat{y}=`r m_bty$coefficients[1]` + `r m_bty$coefficients[2]` x \]
    
    **Interpreting the slope, for every 1 unit increase in beauty index, the course evaluation would increase by ~0.67. While the p-value is less than 0.05, practically speaking, it is not a significant predictor as evidenced by the \(R^2 \approx 0.03\).**

6.  Use residual plots to evaluate whether the conditions of least squares
    regression are reasonable. Provide plots and comments for each one (see
    the Simple Regression Lab for a reminder of how to make these).
    
```{r residuals, eval=TRUE}
plot(m_bty$residuals ~ evals$bty_avg)
abline(h = 0, lty = 3)  # adds a horizontal dashed line at y = 0
```
 
*Nearly normal residuals*: Based on the plot above, the residuals appear to be nearly normal.

*Linearity*: The data are clearly not having a narrow linear relationship. The points are wide in their trend.

## Multiple linear regression

The data set contains several variables on the beauty score of the professor: 
individual ratings from each of the six students who were asked to score the 
physical appearance of the professors and the average of these six scores. Let's 
take a look at the relationship between one of these scores and the average 
beauty score.

```{r bty-rel, eval = TRUE}
plot(evals$bty_avg ~ evals$bty_f1lower)
cor(evals$bty_avg, evals$bty_f1lower)
```

As expected the relationship is quite strong - after all, the average score is 
calculated using the individual scores. We can actually take a look at the 
relationships between all beauty variables (columns 13 through 19) using the 
following command:

```{r bty-rels, eval = TRUE}
plot(evals[,13:19])
```

These variables are collinear (correlated), and adding more than one of these 
variables to the model would not add much value to the model. In this 
application and with these highly-correlated predictors, it is reasonable to use
the average beauty score as the single representative of these variables.

In order to see if beauty is still a significant predictor of professor score 
after we've accounted for the gender of the professor, we can add the gender 
term into the model.

```{r scatter-score-bty_avg_gender, eval = TRUE}
m_bty_gen <- lm(score ~ bty_avg + gender, data = evals, y=TRUE)
summary(m_bty_gen)
```

7.  P-values and parameter estimates should only be trusted if the
    conditions for the regression are reasonable. Verify that the conditions
    for this model are reasonable using diagnostic plots.
    
    **Nearly normal residuals seem to hold as shown in the QQ plot below.**
    
```{r, echo=FALSE}
qqnorm(m_bty_gen$residuals)
qqline(m_bty_gen$residuals) 
```

**The constant variability of residuals is checked via a comparison to the fitted values. As shown in the following visualization, the data appears to satisfy the conditions for regression.**

```{r, echo=FALSE}
dfResVsFit <- data.frame(residuals=abs(residuals(m_bty_gen)), fittedValue=fitted(m_bty_gen))
g7b <- ggplot(data=dfResVsFit) +
  geom_point(aes(x=fittedValue, y=residuals)) + 
  myTheme + 
  labs(title="Residuals vs Fitted Values")
g7b
```

**Based on code from http://www.statmethods.net/stats/regression.html, we can view a full set of diagnostic plots to verify the conditions:**

```{r, fig.width=7.5, fig.height=5, echo=FALSE}
# diagnostic plots 
layout(matrix(c(1,2,3,4),2,2)) # optional 4 graphs/page 
plot(m_bty_gen)
```

**When comparing the residuals to the `bty_avg` predictor values, we see there is not specific structure and the residuals are mostly having constant variability. This supports the concept that the residuals are independent.**

```{r, echo=FALSE}
dfResVsFit <- data.frame(residuals=abs(residuals(m_bty_gen)), bty_avg=evals$bty_avg)
g7c <- ggplot(data=dfResVsFit) +
  geom_point(aes(x=bty_avg, y=residuals)) + 
  myTheme + 
  labs(title="Residuals vs Bty_avg Values")
g7c
```

**Comparing to the `gender` variable, the distribution is slightly different but similar. The difference doesn't appear significant enough to invalidate the model.**

```{r, echo=FALSE}
dfResVsFit <- data.frame(residuals=abs(residuals(m_bty_gen)), gender=evals$gender)
g7d <- ggplot(data=dfResVsFit) +
  geom_boxplot(aes(x=gender, y=residuals)) + 
  myTheme + 
  labs(title="Residuals vs Gender Values")
g7d
```

8.  Is `bty_avg` still a significant predictor of `score`? Has the addition
    of `gender` to the model changed the parameter estimate for `bty_avg`?
    
    **Revisiting the summary of the model we see the parameter estimte for `bty_avg` is now 0.07416 whereas before gender was introduced, the estimate was 0.06664, a `r 0.07416 - 0.06664` increase. Although the parameter estimate changed for `bty_avg`, the t-value is effectively identical and stronger than gender. Seperately, the \(R^2\) is 0.05912 with both `bty_avg` and `gender`, and with only `bty_avg` the \(R^2\) is 0.03502.  Therefore I conclude that adding `gender` to the model is useful, but on the whole, `bty_avg` is not a significant predictor of `score`.**
    
```{r, echo=FALSE}
summary(m_bty_gen)
```

Note that the estimate for `gender` is now called `gendermale`. You'll see this 
name change whenever you introduce a categorical variable. The reason is that R 
recodes `gender` from having the values of `female` and `male` to being an 
indicator variable called `gendermale` that takes a value of $0$ for females and
a value of $1$ for males. (Such variables are often referred to as "dummy" 
variables.)

As a result, for females, the parameter estimate is multiplied by zero, leaving 
the intercept and slope form familiar from simple regression.

\[
  \begin{aligned}
\widehat{score} &= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg + \hat{\beta}_2 \times (0) \\
&= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg\end{aligned}
\]

We can plot this line and the line corresponding to males with the following 
custom function.

```{r twoLines, eval = TRUE}
multiLines(m_bty_gen)
```

9.  What is the equation of the line corresponding to males? (*Hint:* For
    males, the parameter estimate is multiplied by 1.) For two professors
    who received the same beauty rating, which gender tends to have the
    higher course evaluation score?
    
    \[\widehat{score} = `r m_bty_gen$coefficients[1]` + `r m_bty_gen$coefficients[2]` \times bty\_avg + `r m_bty_gen$coefficients[2]` \times (1)\]
    
    **Since we are adding `r m_bty_gen$coefficients[2]` for men, this means men will tend to have a higher course evaluation score. This is evident in the chart shown above also, where the blue line (male) is above the brown line (female).**

The decision to call the indicator variable `gendermale` instead of`genderfemale`
has no deeper meaning. R simply codes the category that comes first 
alphabetically as a $0$. (You can change the reference level of a categorical 
variable, which is the level that is coded as a 0, using the`relevel` function. 
Use `?relevel` to learn more.)

10. Create a new model called `m_bty_rank` with `gender` removed and `rank` 
    added in. How does R appear to handle categorical variables that have more 
    than two levels? Note that the rank variable has three levels: `teaching`, 
    `tenure track`, `tenured`.
    
    **R appears to handle categorical variabels having more than two levels by adding an additional parameter for each additional level above 2. Therefore `teaching` would yield \(\hat{B_2}=0, \hat{B_3}=0\). `tenure track` would be \(\hat{B_2}=1, \hat{B_3}=0\), and `tenured would be \(\hat{B_2}=0, \hat{B_3}=1\) **
    
```{r}
m_bty_rank <- lm(score ~ bty_avg + rank, data=evals)
summary(m_bty_rank)
```

The interpretation of the coefficients in multiple regression is slightly 
different from that of simple regression. The estimate for `bty_avg` reflects
how much higher a group of professors is expected to score if they have a beauty
rating that is one point higher *while holding all other variables constant*. In
this case, that translates into considering only professors of the same rank 
with `bty_avg` scores that are one point apart.

## The search for the best model

We will start with a full model that predicts professor score based on rank, 
ethnicity, gender, language of the university where they got their degree, age, 
proportion of students that filled out evaluations, class size, course level, 
number of professors, number of credits, average beauty rating, outfit, and 
picture color.

11. Which variable would you expect to have the highest p-value in this model? 
    Why? *Hint:* Think about which variable would you expect to not have any 
    association with the professor score.

**I would hypothesize that `language`, language of the university where they got their degree, would
have the highest p-value.**

Let's run the model...

```{r m_full, eval = TRUE, tidy = FALSE}
m_full <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_profs + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full)
```

12. Check your suspicions from the previous exercise. Include the model output
    in your response.
    
    **It appears that cls_prof, number of professors, has the least impact on score with a p-value of 0.77806.**

13. Interpret the coefficient associated with the ethnicity variable.

**Holding all else equal, when ethnicity is `not minority` the score tends to increase by 0.1234929.**

14. Drop the variable with the highest p-value and re-fit the model. Did the
    coefficients and significance of the other explanatory variables change?
    (One of the things that makes multiple regression interesting is that
    coefficient estimates depend on the other variables that are included in
    the model.) If not, what does this say about whether or not the dropped
    variable was collinear with the other explanatory variables?

```{r m_full_less_profs, eval = TRUE, tidy = FALSE}
m_full_less_profs <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full_less_profs)
```

**Coefficients and p-values changed, through only slightly, which suggests some collinearity between the `cls_profs` and other variables.**

15. Using backward-selection and p-value as the selection criterion,
    determine the best model. You do not need to show all steps in your
    answer, just the output for the final model. Also, write out the linear
    model for predicting score based on the final model you settle on.
    
```{r m_backSel_pval, eval = TRUE, tidy = FALSE}
m_backSel_pval <- lm(score ~ ethnicity + gender + language + age + cls_perc_eval 
             + cls_credits + bty_avg 
             + pic_color, data = evals)
summary(m_backSel_pval)

coeffBackSel <- coefficients(m_backSel_pval)
```

\[
  \begin{aligned}
  \widehat{score} = & `r coeffBackSel[1]` + `r coeffBackSel[2]` \text{ ethnicity}_{not minority} + `r coeffBackSel[3]`  \text{ gender}_{male} + \\ 
  & `r coeffBackSel[4]` \text{ language}_{non-english} + `r coeffBackSel[5]` \text{ age} + `r coeffBackSel[6]` \text{ cls\_perc\_eval } + \\ 
  & `r coeffBackSel[7]` \text{ cls\_credits}_{one credit} + `r coeffBackSel[8]`  \text{ bty\_avg} + `r coeffBackSel[9]` \text{ pic\_color}_{color}\end{aligned}
\]

16. Verify that the conditions for this model are reasonable using diagnostic 
    plots.
    
**Normalness appears fine:**

```{r, echo=FALSE}
qqnorm(m_backSel_pval$residuals)
qqline(m_backSel_pval$residuals) 
```

**The constant variability of residuals is checked via a comparison to the fitted values. As shown in the following visualization, the data appears to satisfy the conditions for regression.**

```{r, echo=FALSE}
dfResVsFit <- data.frame(residuals=abs(residuals(m_backSel_pval)), fittedValue=fitted(m_backSel_pval))
g7b <- ggplot(data=dfResVsFit) +
  geom_point(aes(x=fittedValue, y=residuals)) + 
  myTheme + 
  labs(title="Residuals vs Fitted Values")
g7b
```


17. The original paper describes how these data were gathered by taking a
    sample of professors from the University of Texas at Austin and including 
    all courses that they have taught. Considering that each row represents a 
    course, could this new information have an impact on any of the conditions 
    of linear regression?
    
    **The new information (all courses they have taught) would introduce bias toward professors who have been at the universtiy longer whom would be represented more in sample.**

18. Based on your final model, describe the characteristics of a professor and 
    course at University of Texas at Austin that would be associated with a high
    evaluation score.
    
```{r}
#summary(evals)
# Characteristics of a high scoring professor
vHigh <- c(1,     # Intercept
           1,     # Ethnicity: Non minority
           1,     # Gender: Male
           0,     # Language: Non-english
           29,    # Age 
           100,   # Percent of Class Evaluations 
           1,     # Credits: One Credit
           8.167, # Average Beauty Score
           0)     # Pic Color: Color
# Compute predicted score
score <- coefficients(m_backSel_pval) %*% vHigh
# Exceeds the maximum
score
```

19. Would you be comfortable generalizing your conclusions to apply to professors
    generally (at any university)? Why or why not?
    
    **I would not be comfortable generalizing these conclusions to professors at other univesities, but could be used more generally at the University of Texas Austin. The population being sampled is specifically the UoTA professors. Inorder to generalize to any university, we would need to sample from all unversities in some way and would likely include a geographic factor in the model as well.**

<div id="license">
This is a product of OpenIntro that is released under a [Creative Commons Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0). This lab was written by 
Mine &Ccedil;etinkaya-Rundel and Andrew Bray.
</div>
