---
title: "IS606 Data Project Proposal"
author: "Daniel Dittenhafer"
date: "October 19, 2015"
output: pdf_document
---
```{r include=FALSE, cache=FALSE}
# DO NOT REMOVE
# THIS IS FOR SETTING SOME PLOTTING PARAMETERS SO THAT YOUR PLOTS DON'T TAKE UP TOO MUCH SPACE
# IF YOU WOULD LIKE TO CHANGE THESE, SEE HELP FILES ON THE par() FUNCTION
# OR ASK FOR HELP
library(knitr)
## set global chunk options
opts_chunk$set(cache.path='cache/manual-', fig.show='hold', fig.path='figure/manual-', par=TRUE)# fig.width=4, fig.height=3, fig.align='center')
## tune details of base graphics (http://yihui.name/knitr/hooks)
knit_hooks$set(par=function(before, options, envir){
if (before && options$fig.show!='none') par(mar=c(4,4,.2,.1),cex.lab=.95,cex.axis=.9,mgp=c(2,.7,0),tcl=-.3)
})
```

```{r, echo=FALSE}
library(plyr)
library(ggplot2)
library(knitcitations)
library(RefManageR)
cleanbib()
cite_options(style="markdown")

bibWorldValSurvey <- bibentry(bibtype="Misc", 
                              author=person(family="World Values Survey Association"),
                              title="WORLD VALUES SURVEY 1981-2014 LONGITUDINAL AGGREGATE v.20150418",
                              year=2014,
                              url="http://www.worldvaluessurvey.org/WVSDocumentationWVL.jsp",
                              note="Aggregate File Producer: JDSystems",
                              address="Madrid SPAIN"
                              )

bibWvsFieldworkSampling <- bibentry(bibtype="Misc", 
                              author=person(family="World Values Survey Association"),
                              title="Fieldwork and Sampling",
                              year=2015,
                              url="http://www.worldvaluessurvey.org/WVSContents.jsp")

myTheme <- theme(axis.ticks=element_blank(),  
                  panel.border = element_rect(color="gray", fill=NA), 
                  panel.background=element_rect(fill="#FBFBFB"), 
                  panel.grid.major.y=element_line(color="white", size=0.5), 
                  panel.grid.major.x=element_line(color="white", size=0.5),
                  plot.title=element_text(size="10"))

# 
# Original WVS Data File, Size 1.4GB
#
#dataFile <- "C:/Users/Dan/Downloads/WorldValuesSurvey/WVS_Longitudinal_1981-2014_rdata_v_2015_04_18/WVS_Longitudinal_1981_2014_R_v2015_04_18.rdata"
#load(dataFile)
#colsToKeep <- c("E234", "S020",  "S024", "S003", "X003", "X025", "A006")
#WVS_Subset <- WVS_Longitudinal_1981_2014_R_v2015_04_18[,colsToKeep]
#save(WVS_Subset, file="C:/Code/R/IS606-ProbStats/FinalProject/data/WVS_subData.RData")
# My project: C:/Code/R/IS606-ProbStats/FinalProject
```

```{r}
# load data
dataFile <- "data/WVS_subData.RData"
load(dataFile)
colnames(WVS_Subset) <- c("KeyQuestion",
                          "Year", 
                          "CountryWave", 
                          "Country", 
                          "Age", 
                          "HighestEducation",
                          "ReligionImportant")
# Subset to just United States
WVS_US <- WVS_Subset[WVS_Subset$Country == 840, ]
WVS_US$CountryName <- "United States"
```


### Research question 

**You should phrase your research question in a way that matches up with the scope of inference your dataset allows for.**

How have the values of people from the United States changed over time with regard to science and technology and its positive/negative impact on the world? Is there a significant difference in views at different educational levels? If time allows, the importance of religion will also be investigated.

### Cases 

**What are the cases, and how many are there?**

Each row in the data set is a case and represents the results of an interview conducted by the World Values Survey with an individual age 18 or above. `r citep(bibWvsFieldworkSampling)`

There are `r nrow(WVS_US)` cases in the subset of data to be used in the project.
```{r}
nrow(WVS_US)
```

### Data collection 

**Describe the method of data collection.**

According to the World Values Survey website, the data is collected either through "face-to-face interviews or phone interviews for remote areas." `r citep(bibWvsFieldworkSampling)`

### Type of study 

**What type of study is this (observational/experiment)?**

The World Values Survey is an observational study, and this data project will be analyzing a subset of the observational study conducted by the World Values Survey Association.


### Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**

The data was collected and made available by the World Values Survey Association. `r  citep(bibWorldValSurvey)`.

The link to the data page is: http://www.worldvaluessurvey.org/WVSDocumentationWVL.jsp

#### Variables Included in Data Subset

The following table lists the variables extracted from the original World Values Survey dataset which will be considered in this data project.

  | Variables | Description
--|-----------|------------
005_203| E234 | The world is better off, or worse off, because of science and technology
010_023| S020 | Year survey
010_028| S024 | Country wave
010_004| S003 | Country/region
014_003| X003 | Age
014_030| X025 |	Highest educational level attained
001_006| A006	| Important in life: Religion


### Response 

**What is the response variable, and what type is it (numerical/categorical)?**

The response variable is the answer provided to the key question, "Is the world is better off, or worse off, because of science and technology?"

The answers are categorical in nature, but ordinal in their degree of support for the better/worse outcome. The distinct answer values and the description of each value are shown in the table below.

Value | Description
------|------------
1|A lot worse off
2|2
3|3
4|4
5|5
6|6
7|7
8|8
9|9
10|A lot better off
-5|Missing; Unknown
-4|Not asked in survey
-3|Not applicable
-2|No answer
-1|Don´t know

### Explanatory 

**What is the explanatory variable, and what type is it (numerical/categorical)?**

The explanatory variables I've choosen to analyze are the "Highest educational level attained" value and the "Important in life: Religion" variable. Again, these are categorical variables, but ordinal in terms of level of education and degree of importance of religion. The values and descriptions for each variable follow:

**Highest educational level attained**

Value | Description
------|------------
1|Inadequately completed elementary education
2|Completed (compulsory) elementary education
3|Incomplete secondary school: technical/vocational type/elementary education, basic vocational qual.
4|Complete secondary school: technical/vocational type/Secondary, intermediate vocational qualification
5|Incomplete secondary: university-preparatory type/Secondary, intermediate general qualification
6|Complete secondary: university-preparatory type/Full secondary, maturity level certificate
7|Some university without degree/Higher education - lower-level tertiary certificate
8|University with degree/Higher education - upper-level tertiary certificate
-5|Missing; Unknown
-4|Not asked in survey
-3|Not applicable; No formal education
-2|No answer
-1|Don´t know

**Important in life: Religion**

Value | Description
------|------------
-5|Missing; Unknown
-4|Not asked in survey
-3|Not applicable
-2|No answer
-1|Don´t know
1|Very important
2|Rather important
3|Not very important
4|Not at all important


### Relevant summary statistics 

**Provide summary statistics relevant to your research question. For example, if you're comparing means across groups provide means, SDs, sample sizes of each group. This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

First, we show raw summary statistics for the data points:

```{r, echo=FALSE}
summary(WVS_US)
```

The following `R` output shows the mean and standard deviation of the response variable for the various years for which the survey was conducted. Unfortunately, in 1995 and 1999 the question regarding science was not asked as evidenced by the mean of -4 and standard deviation of 0. As a result, the focus will be on differences between 2006 and 2011.

```{r, echo=FALSE}
plyr::ddply(WVS_US, .(Year), summarize, 
            meanResponse=mean(KeyQuestion), 
            sdResponse=sd(KeyQuestion))

WVS_US_9599 <- WVS_US[WVS_US$Year > 1999,]
```

After eliminating the 1995 and 1999 data rows, `r nrow(WVS_US_9599)` cases remain.

The following `R` output shows the mean and standard deviation at each level of the response variable for the education and religion explanatory variables for these remaining cases from 2006 and 2011.

```{r, echo=FALSE}
plyr::ddply(WVS_US_9599, .(KeyQuestion), summarize, 
            meanEducation=mean(HighestEducation), 
            sdEducation=sd(HighestEducation), 
            meanReligion=mean(ReligionImportant),
            sdReligion=sd(ReligionImportant))
```

```{r, data-distrib-year, echo=FALSE, message=FALSE}
g1 <- ggplot(data=WVS_US_9599) + 
  geom_bar(aes(x=Year), binwidth=1) +
  myTheme +
  labs(title="Response Distribution By Year")
g1
```

### References

```{r, results='asis', echo=FALSE}
BibOptions(style="html", bib.style="authortitle")
bibliography()
```
