---
title: "IS606 Data Project"
author: "Daniel Dittenhafer"
date: "September 28, 2015"
output: pdf_document
---
```{r include=FALSE, cache=FALSE}
# DO NOT REMOVE
# THIS IS FOR SETTING SOME PLOTTING PARAMETERS SO THAT YOUR PLOTS DON'T TAKE UP TOO MUCH SPACE
# IF YOU WOULD LIKE TO CHANGE THESE, SEE HELP FILES ON THE par() FUNCTION
# OR ASK FOR HELP
library(knitr)
## set global chunk options
opts_chunk$set(fig.path='figure/manual-', cache.path='cache/manual-', fig.align='center', fig.show='hold', par=TRUE)
## tune details of base graphics (http://yihui.name/knitr/hooks)
knit_hooks$set(par=function(before, options, envir){
if (before && options$fig.show!='none') par(mar=c(4,4,.2,.1),cex.lab=.95,cex.axis=.9,mgp=c(2,.7,0),tcl=-.3)
})
```

```{r}
# load data
#dataFile <- "C:/Users/Dan/Downloads/WorldValuesSurvey/WVS_Longitudinal_1981-2014_rdata_v_2015_04_18/WVS_Longitudinal_1981_2014_R_v2015_04_18.rdata"
dataFile <- "C:/Code/R/IS606-ProbStats/FinalProject/data/WVS_subData.RData"
load(dataFile)

summary(WVS_Subset)

```


### Research question 

**You should phrase your research question in a way that matches up with the scope of inference your dataset allows for.**

How have the values of people from the United States changed over time with regard to science and technology and its positive/negative impact on the world.

  | Variables | Description
--|-----------|------------
005_203| E234 | The world is better off, or worse off, because of science and technology
010_023| S020 | Year survey
010_028| S024 | Country wave
010_004| S003 | Country/region
014_003| X003 | Age
014_030| X025 |	Highest educational level attained


### Cases 

**What are the cases, and how many are there?**



### Data collection 

**Describe the method of data collection.**



### Type of study 

**What type of study is this (observational/experiment)?**



### Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**




### Response 

**What is the response variable, and what type is it (numerical/categorical)?**



### Explanatory 

**What is the explanatory variable, and what type is it (numerical/categorival)?**



### Relevant summary statistics 

**Provide summary statistics relevant to your research question. For example, if you’re comparing means across groups provide means, SDs, sample sizes of each group. This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r}

```


