---
title: "Introduction to linear regression"
output:
  pdf_document: default
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
---
```{r, echo=FALSE} 
library(ggplot2)
myTheme <- theme(axis.ticks=element_blank(),  
                  panel.border = element_rect(color="gray", fill=NA), 
                  panel.background=element_rect(fill="#FBFBFB"), 
                  panel.grid.major.y=element_line(color="white", size=0.5), 
                  panel.grid.major.x=element_line(color="white", size=0.5),
                  plot.title=element_text(size="10"))
```

## Batter up 

The movie [Moneyball](http://en.wikipedia.org/wiki/Moneyball_(film)) focuses on
the "quest for the secret of success in baseball". It follows a low-budget team, 
the Oakland Athletics, who believed that underused statistics, such as a player's 
ability to get on base, betterpredict the ability to score runs than typical 
statistics like home runs, RBIs (runs batted in), and batting average. Obtaining 
players who excelled in these underused statistics turned out to be much more 
affordable for the team.

In this lab we'll be looking at data from all 30 Major League Baseball teams and
examining the linear relationship between runs scored in a season and a number 
of other player statistics. Our aim will be to summarize these relationships 
both graphically and numerically in order to find which variable, if any, helps 
us best predict a team's runs scored in a season.

## The data

Let's load up the data for the 2011 season.

```{r load-data, eval=TRUE}
load("more/mlb11.RData")
```

In addition to runs scored, there are seven traditionally used variables in the 
data set: at-bats, hits, home runs, batting average, strikeouts, stolen bases, 
and wins. There are also three newer variables: on-base percentage, slugging 
percentage, and on-base plus slugging. For the first portion of the analysis 
we'll consider the seven traditional variables. At the end of the lab, you'll 
work with the newer variables on your own.

1.  What type of plot would you use to display the relationship between `runs` 
    and one of the other numerical variables? Plot this relationship using the 
    variable `at_bats` as the predictor. Does the relationship look linear? If 
    you knew a team's `at_bats`, would you be comfortable using a linear model 
    to predict the number of runs?
    
    **Based on the visualization below, there does seem to be a positive relationship between `at_bats` and `runs`. The spread is fairly wide, so a prediction might not be very useful alone, but it appears strong enough to get a flavor.**
    
```{r}
g1 <- ggplot(data=mlb11, aes(x=at_bats, y=runs)) + geom_point() + myTheme
g1
```

If the relationship looks linear, we can quantify the strength of the
relationship with the correlation coefficient.

```{r cor, eval=TRUE}
cor(mlb11$runs, mlb11$at_bats)
```

## Sum of squared residuals

Think back to the way that we described the distribution of a single variable. 
Recall that we discussed characteristics such as center, spread, and shape. It's
also useful to be able to describe the relationship of two numerical variables, 
such as `runs` and `at_bats` above.

2.  Looking at your plot from the previous exercise, describe the relationship 
    between these two variables. Make sure to discuss the form, direction, and 
    strength of the relationship as well as any unusual observations.
    
    **There is a positive relationship. As `at_bats` increase, `runs` tend to also. This is not an absolute relationship, nor a tight extremely linear one. None the less, there is a trend in the data from lower left to upper right.**

Just as we used the mean and standard deviation to summarize a single variable, 
we can summarize the relationship between these two variables by finding the 
line that best follows their association. Use the following interactive 
function to select the line that you think does the best job of going through 
the cloud of points.

```{r plotss-atbats-runs, eval=TRUE}
plot_ss(x = mlb11$at_bats, y = mlb11$runs)
```

After running this command, you'll be prompted to click two points on the plot 
to define a line. Once you've done that, the line you specified will be shown in
black and the residuals in blue. Note that there are 30 residuals, one for each 
of the 30 observations. Recall that the residuals are the difference between the
observed values and the values predicted by the line:

\[
  e_i = y_i - \hat{y}_i
\]

The most common way to do linear regression is to select the line that minimizes
the sum of squared residuals. To visualize the squared residuals, you can rerun 
the plot command and add the argument `showSquares = TRUE`.

```{r plotss-atbats-runs-squares, eval=FALSE}
plot_ss(x = mlb11$at_bats, y = mlb11$runs, showSquares = TRUE)
```

Note that the output from the `plot_ss` function provides you with the slope and
intercept of your line as well as the sum of squares.

3.  Using `plot_ss`, choose a line that does a good job of minimizing the sum of
    squares. Run the function several times. What was the smallest sum of 
    squares that you got? How does it compare to your neighbors?
    
    **Smallest sum of squares was 145516.1. Largest was 183082.8**
    
    

## The linear model

It is rather cumbersome to try to get the correct least squares line, i.e. the 
line that minimizes the sum of squared residuals, through trial and error. 
Instead we can use the `lm` function in R to fit the linear model (a.k.a. 
regression line).

```{r m1, eval=TRUE}
m1 <- lm(runs ~ at_bats, data = mlb11)
```

The first argument in the function `lm` is a formula that takes the form 
`y ~ x`. Here it can be read that we want to make a linear model of `runs` as a 
function of `at_bats`. The second argument specifies that R should look in the 
`mlb11` data frame to find the `runs` and `at_bats` variables.

The output of `lm` is an object that contains all of the information we need 
about the linear model that was just fit. We can access this information using 
the summary function.

```{r summary-m1, eval=TRUE}
summary(m1)
```

Let's consider this output piece by piece. First, the formula used to describe 
the model is shown at the top. After the formula you find the five-number 
summary of the residuals. The "Coefficients" table shown next is key; its first 
column displays the linear model's y-intercept and the coefficient of `at_bats`.
With this table, we can write down the least squares regression line for the 
linear model:

\[
  \hat{y} = -2789.2429 + 0.6305 * atbats
\]

One last piece of information we will discuss from the summary output is the 
Multiple R-squared, or more simply, $R^2$. The $R^2$ value represents the 
proportion of variability in the response variable that is explained by the 
explanatory variable. For this model, 37.3% of the variability in runs is 
explained by at-bats.

4.  Fit a new model that uses `homeruns` to predict `runs`. Using the estimates 
    from the R output, write the equation of the regression line. What does the 
    slope tell us in the context of the relationship between success of a team 
    and its home runs?
    
```{r}
m2 <- lm(runs ~ homeruns, data = mlb11)
summary(m2)
```

\[
  \hat{y} = `r m2$coefficients[1]` + `r m2$coefficients[2]` * homeruns
\]

**The slope of `r m2$coefficients[2]` tells us there is a `r m2$coefficients[2]` run increase for every 1 homerun increase. As compare to the `at_bats` slope, the homeruns slope is steeper.**

## Prediction and prediction errors

Let's create a scatterplot with the least squares line laid on top.

```{r reg-with-line, eval=TRUE}
plot(mlb11$runs ~ mlb11$at_bats)
abline(m1)
```

The function `abline` plots a line based on its slope and intercept. Here, we 
used a shortcut by providing the model `m1`, which contains both parameter 
estimates. This line can be used to predict $y$ at any value of $x$. When 
predictions are made for values of $x$ that are beyond the range of the observed
data, it is referred to as *extrapolation* and is not usually recommended. 
However, predictions made within the range of the data are more reliable. 
They're also used to compute the residuals.

5.  If a team manager saw the least squares regression line and not the actual 
    data, how many runs would he or she predict for a team with 5,578 at-bats? 
    Is this an overestimate or an underestimate, and by how much? In other 
    words, what is the residual for this prediction?
    
```{r}
atbats <- 5578 
prRuns <- −2789.2429 + 0.6305 ∗ atbats
prRuns

mlb11[mlb11$atbats == atbats,]
```

**Due to the prediction being based on and possibly exactly on the regression line, the residual would zero or close to it since the residual is the different between a data point and the regression line. In terms of over or underestimating, the regression line is a little high and a little low in this range. depending on which side of the 5,578 point we want to use to measure the estimate. **

## Model diagnostics

To assess whether the linear model is reliable, we need to check for (1) 
linearity, (2) nearly normal residuals, and (3) constant variability.

*Linearity*: You already checked if the relationship between runs and at-bats
is linear using a scatterplot. We should also verify this condition with a plot 
of the residuals vs. at-bats. Recall that any code following a *#* is intended
to be a comment that helps understand the code but is ignored by R.

```{r residuals, eval=TRUE}
plot(m1$residuals ~ mlb11$at_bats)
abline(h = 0, lty = 3)  # adds a horizontal dashed line at y = 0
```

6.  Is there any apparent pattern in the residuals plot? What does this indicate
    about the linearity of the relationship between runs and at-bats?
    
    **There seems to be more points on the left versus the right side, but generally constant variability with a similar amount of point above and below the line. This suggests a somewhat linear though wide relationship.**

*Nearly normal residuals*: To check this condition, we can look at a histogram

```{r hist-res, eval=TRUE}
hist(m1$residuals)
```

or a normal probability plot of the residuals.

```{r qq-res, eval=TRUE}
qqnorm(m1$residuals)
qqline(m1$residuals)  # adds diagonal line to the normal prob plot
```

7.  Based on the histogram and the normal probability plot, does the nearly 
    normal residuals condition appear to be met?
    
    **I would say the distribution is not nearly as normal as we would like, though there is a weak normal shape.**

*Constant variability*:

8.  Based on the plot in (1), does the constant variability condition appear to 
    be met?
    
    **It appears the constant variability condition would be satisfied.**
    
* * *

## On Your Own

-   Choose another traditional variable from `mlb11` that you think might be a 
    good predictor of `runs`. Produce a scatterplot of the two variables and fit 
    a linear model. At a glance, does there seem to be a linear relationship?

**There does _not_ seem to be a linear relationship between `stolen_bases` and `runs`.**

```{r}
g3 <- ggplot(data=mlb11, aes(x=stolen_bases, y=runs)) + geom_point() + myTheme
```

```{r}
m3 <- lm(runs ~ stolen_bases, data=mlb11)
g3 + geom_abline(intercept=m3$coefficients[1], slope=m3$coefficients)
summary(m3)
```

-   How does this relationship compare to the relationship between `runs` and 
    `at_bats`? Use the R$^2$ values from the two model summaries to compare. 
    Does your variable seem to predict `runs` better than `at_bats`? How can you
    tell?
    
**The `stolen_bases` \(R^2=0.002914\) which is much lower than the `at_bats` \(R^2=0.3729\). This indicates that the `at_bats` model explains significantly more of the variability of `runs` than `stolen_bases` does.**

-   Now that you can summarize the linear relationship between two variables, 
    investigate the relationships between `runs` and each of the other five 
    traditional variables. Which variable best predicts `runs`? Support your 
    conclusion using the graphical and numerical methods we've discussed (for 
    the sake of conciseness, only include output for the best variable, not all 
    five).
    
    **Based on a correlation analysis, `bat_avg` appears to have the best positive relationship with `runs` at ~0.8099. The scatterplot below shows a general linear positive trend in the data.**
    
```{r}
g4 <- ggplot(data=mlb11, aes(x=bat_avg, y=runs)) + geom_point() + myTheme
g4

m4 <- lm(runs ~ bat_avg, data=mlb11)
summary(m4)
```

**The \(R^2=0.6561\), and the model would be as follows:**

\[
  \hat{y} = `r m4$coefficients[1]` + `r m4$coefficients[2]` * bat_avg
\]

The following visualization shows the residuals. No obvious pattern is apparent.

```{r}
resData <- data.frame(residuals=m4$residuals, bat_avg=mlb11$bat_avg)
g5 <- ggplot(data=resData, aes(x=bat_avg, y=residuals)) + 
  geom_point() + 
  geom_hline(yintercept=0) +
  myTheme
g5
```

-   Now examine the three newer variables. These are the statistics used by the 
    author of *Moneyball* to predict a teams success. In general, are they more 
    or less effective at predicting runs that the old variables? Explain using 
    appropriate graphical and numerical evidence. Of all ten variables we've 
    analyzed, which seems to be the best predictor of `runs`? Using the limited 
    (or not so limited) information you know about these baseball statistics, 
    does your result make sense?
    
    **As shown below, the correlation coefficents are higher, in the low to mid 90s, versus the low 80s with the best standard statistic `bat_avg`. The visualizations that follow show a much stronger linear relationship with a narrower range across the data. This supports a better predictive outcome of a linear regression model. ** 
    
```{r}
cor(mlb11[,c(2,10,11,12)])
```

```{r}
g6 <- ggplot(data=mlb11, aes(x=new_onbase, y=runs)) + geom_point() + myTheme
g6
```


```{r}
g6 <- ggplot(data=mlb11, aes(x=new_slug, y=runs)) + geom_point() + myTheme
g6
```

```{r}
g6 <- ggplot(data=mlb11, aes(x=new_obs, y=runs)) + geom_point() + myTheme
g6
```

**Moving forward with `new_obs`, let's check the linear model. We see from the summary output that the \(R^2=0.9349\). I'm not sure what the `new_obs` represents, but it seems to be a good predictor. P-values are well below 0.05.**

```{r}
m6 <- lm(runs ~ new_obs, data=mlb11)
summary(m6)
```

The residual plot below looks good. No obvious trends, and constant variability.

```{r}
resData <- data.frame(residuals=m6$residuals, new_obs=mlb11$new_obs)
g5 <- ggplot(data=resData, aes(x=new_obs, y=residuals)) + 
  geom_point() + 
  geom_hline(yintercept=0) +
  myTheme
g5
```

-   Check the model diagnostics for the regression model with the variable you 
    decided was the best predictor for runs.
    
    **P-values are very low, well below 0.05, and \(R^2\) is high, at 0.9349, so `new_obs` seems like a good predictor of runs.**

<div id="license">
This is a product of OpenIntro that is released under a [Creative Commons 
Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0). 
This lab was adapted for OpenIntro by Andrew Bray and Mine &Ccedil;etinkaya-Rundel 
from a lab written by the faculty and TAs of UCLA Statistics.
</div>
